---
title: 'Realistic populations: age pyramids and dates of death'
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, message = FALSE, warning = FALSE)
library(razer)
set_seed(1)   # seeds razer's internal RNG (used by the alias sampler and Kaplan–Meier)
```

> Companion to `examples/aged_population.R`.

## Why bother with demographics?

Many questions need agents that *aren't* interchangeable: an SIA that targets under-fives,
age-dependent severity, vital dynamics where newborns enter susceptible and the old die.
That requires two ingredients at initialization — a realistic **age structure**, and a
**date of death** for every agent (so natural mortality can simply retire agents when their
clock arrives). razer provides both, ported from `laser.core`. No disease here; this is the
population-building step that the vital-dynamics and measles models build on.

## Ages from a population pyramid (the alias method)

A *population pyramid* is the count of people in each age band (often by sex). To sample a
million agents' ages in proportion to it, razer builds an **`AliasedDistribution`** — Walker's
alias method, which draws from an arbitrary discrete distribution in **O(1)** per sample
after an O(k) setup, far faster than a cumulative-sum search. `sample_pyramid_ages()` wraps
it: pick a band in proportion to its population, then a uniform day within that band's year
range, returning ages in **days**.

```{r ages}
pyramid    <- load_pyramid_csv(file.path("..", "data", "pyramid_example.csv"))  # "Age,M,F" bands
n_agents   <- 1000000L
ages_days  <- sample_pyramid_ages(pyramid, n_agents)
ages_years <- ages_days / 365
cat(sprintf("%s agents; ages %.1f–%.1f yr (mean %.1f)\n",
            format(n_agents, big.mark = ","), min(ages_years), max(ages_years), mean(ages_years)))
```

## A date of death (Kaplan–Meier, conditioned on current age)

We want each agent's *age at death*. razer's **`KaplanMeierEstimator`** takes a life table —
**cumulative deaths by year** for a synthetic cohort — and samples an age at death for each
agent **conditioned on the age it is alive at now**, so the draw is never in the past (the
essence of Kaplan–Meier conditional survival). Here we synthesize a life table from a
Gompertz–Makeham per-year mortality hazard $q(a)$ — a small baseline, an exponential rise
with age, and an elevated infant rate — by running a cohort through it.

```{r lifetable}
n_years <- 106L; age <- 0:(n_years - 1L)
q <- pmin(0.0004 + 1e-5 * exp(0.115 * age), 1); q[1] <- 0.02      # hazard; q[1] = infant mortality
deaths <- numeric(n_years); alive <- 1e5
for (a in seq_len(n_years)) {                                    # run a cohort through the hazard
  d <- if (a == n_years) alive else round(alive * q[a])          # force the last year to clear
  deaths[a] <- d; alive <- alive - d
}
km <- kaplan_meier_estimator(cumsum(deaths))                     # non-decreasing cumulative deaths

death_age_days <- km$predict_age_at_death(ages_days, -1L)        # -1L: use the table's last year
death_age_years <- death_age_days / 365
stopifnot(all(death_age_days >= ages_days))                      # nobody dies in the past
e0 <- mean(km$predict_age_at_death(integer(1e5L), -1L)) / 365    # life expectancy at birth
cat(sprintf("predicted age at death: mean %.1f yr; implied e0 = %.1f yr; mean remaining life %.1f yr\n",
            mean(death_age_years), e0, mean(death_age_years - ages_years)))
```

```{r plots}
par(mfrow = c(2, 1), mar = c(4, 4.5, 2.5, 1)); br <- seq(0, n_years, 1)
hist(ages_years, breaks = br, col = "steelblue", border = "white",
     xlab = "age (years)", ylab = "agents", main = "Sampled age structure (the pyramid, realized)")
hist(death_age_years, breaks = br, col = "firebrick", border = "white",
     xlab = "age at death (years)", ylab = "agents",
     main = "Predicted age at death (note the infant spike and the old-age bulk)")
abline(v = e0, lty = 2, col = "grey30"); legend("topleft", sprintf("e0 = %.1f yr", e0), lty = 2, bty = "n")
```

## How this feeds a model

Store the results as per-agent Columns and the rest of razer can use them:

- **`dob` (date of birth) = −age in days**, an `i32` Column (negative = born before $t_0$).
- **`dod` (date of death) = an absolute tick**, a `u32` Column; the `mortality()` kernel
  retires an agent when `dod <= tick`. The `births()` kernel draws a fresh `dod` for each
  newborn from the same `km`.
- Age-based **targeting**: `bincount_where(nodeid, n_nodes, dob, "gt", tick - 5*365, count)`
  counts (or, with `_wt`, weights) the under-fives per node for a campaign or a report.

```{r columns}
dob <- allocate_scalar("i32", n_agents); dob$set(-ages_days)
dod <- allocate_scalar("u32", n_agents); dod$set(as.integer(death_age_days))
cat(sprintf("under-5s at t=0: %s of %s agents\n",
            format(sum(dob$values() > -5L * 365L), big.mark = ","), format(n_agents, big.mark = ",")))
```

## Customize and extend

- **Real pyramids / life tables.** Replace `pyramid_example.csv` with a UN/national pyramid,
  and build `km` from a published abridged life table (cumulative deaths by age) instead of
  the synthetic Gompertz hazard.
- **Use it in a model.** Pass these `dob`/`dod` Columns into a `run_model()` `init` callback
  and run the `births`/`mortality` kernels in `step_exit` — see
  `examples/engwal_measles.R` and the vital-dynamics notebook.
- **Sizing the array.** A growing population needs head-room: `calc_capacity()` bounds the
  cumulative births, `calc_capacity_cdr()` the peak-living size for a `squash()`-reclaimed
  run (see the long-runs notebook).
