---
title: "Interventions: custom states, vaccination campaigns, and quarantine"
output:
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, message = FALSE, warning = FALSE)
library(razer)
set.seed(20260608L)
```

> Companion to `examples/sia_campaigns.R`, `examples/sia_campaigns_waning.R`, and
> `examples/quarantine.R`. This notebook is about **extending** razer.

## The extension toolkit

`run_model()` runs the built-in S/E/I/R disease step, but most public-health questions need
*extra* states and *interventions*. razer exposes four hooks so you never have to fork the
runner or write a Rust kernel:

| Tool | What it does |
|---|---|
| `extra_states = "X"` | registers a **new agent state** `X` (a fresh code in `model$states`): a census Column carried forward and totalled into `N`. The disease kernels ignore unknown states, so an agent in `X` is inert — not infected, not transitioned — until *you* move it. |
| `init(model)` | once, before the loop: add per-agent Columns, seed states. |
| `step_enter` / `step_update` / `step_exit` | per-tick callbacks (start / between step & `calc_foi` / end). Each receives the `model` environment (`$people`, `$nodes`, `$states`, `$tick`). |
| `step_timer_expire(state, timer, …, from, to)` | a generic kernel: decrement each agent's `u16` timer and, on expiry, move `from → to`. The building block for *any* timed transition. |
| `move_count(from, to, counts, tick)` | apply a per-node census delta (either side may be `NULL`). |

An intervention is therefore: in a callback, change some agents' `state` (and `timer`), then
`move_count` the matching census delta. Nothing in the kernels changes.

## 1. A vaccination campaign (S → V)

A Supplemental Immunization Activity: on a schedule, in targeted nodes, move each
susceptible to a new `"V"` state with probability `coverage`. `V` is protected — the force
of infection reads only the `I` census, and the disease step never touches `V`. Three
independent patches, with two targeted and one control, make the node-targeting visible.

```{r sia}
n_nodes <- 3L; pop <- 3e5L; nticks <- 3L * 365L
campaign_ticks <- seq(30L, nticks - 2L, by = 365L)   # annual, first one early (preventive)
affected <- c(0L, 1L); coverage <- 0.6

campaign <- function(model) {
  if (!(model$tick %in% campaign_ticks)) return(invisible())
  V <- model$states[["V"]]; S <- model$states[["S"]]
  s <- model$people$state$values(); nid <- model$people$nodeid$values()
  cnt <- model$people$count
  elig <- which(s[seq_len(cnt)] == S & nid[seq_len(cnt)] %in% affected)
  vacc <- elig[runif(length(elig)) < coverage]                # probabilistic coverage
  if (!length(vacc)) return(invisible())
  s[vacc] <- V; model$people$state$set(s)
  move_count(model$nodes$S, model$nodes$V, tabulate(nid[vacc] + 1L, model$nodes$count), model$tick)
}

vax <- run_model(data.frame(population = rep(pop, n_nodes), I = rep(50L, n_nodes)),
                 "SIR", nticks = nticks, r0 = 2, infectious_period = dist_gamma(2, 4),
                 network = matrix(0, n_nodes, n_nodes), seed = 1L,
                 extra_states = "V", step_exit = campaign)

yr <- (seq_len(nticks) - 1L) / 365; col <- c("#1b9e77", "#7570b3", "#d95f02")
par(mfrow = c(2, 1), mar = c(4, 4.5, 2.5, 1))
matplot(yr, vax$nodes$V$values() / 1e3, type = "l", lty = 1, lwd = 2, col = col,
        xlab = "year", ylab = "vaccinated (000s)", main = "V staircases up in the two targeted nodes")
legend("topleft", c("node 0 (targeted)", "node 1 (targeted)", "node 2 (control)"), col = col, lwd = 2, bty = "n")
matplot(yr, vax$nodes$I$values() / 1e3, type = "l", lty = 1, lwd = 2, col = col,
        xlab = "year", ylab = "infectious (000s)", main = "...and only the control burns through")
```

## 2. Add waning immunity (V → S) — one extra callback

Permanent immunity above is just the *absence* of a return transition. To make `V` wane,
set a timer when vaccinating and add a `step_update` that runs `step_timer_expire(V → S)`
each tick. That's the **entire** change — no new kernel, no disease-step edit.

```{r waning}
waning_days <- dist_normal(2 * 365, 90)   # ~2-year immunity, per-agent draw

campaign_timed <- function(model) {        # as before, but also set the waning timer
  if (!(model$tick %in% campaign_ticks)) return(invisible())
  V <- model$states[["V"]]; S <- model$states[["S"]]
  s <- model$people$state$values(); nid <- model$people$nodeid$values(); tm <- model$people$timer$values()
  cnt <- model$people$count
  elig <- which(s[seq_len(cnt)] == S & nid[seq_len(cnt)] %in% affected)
  vacc <- elig[runif(length(elig)) < coverage]
  if (!length(vacc)) return(invisible())
  s[vacc] <- V; tm[vacc] <- pmax(1L, round(waning_days$sample_n(length(vacc))))
  model$people$state$set(s); model$people$timer$set(tm)
  move_count(model$nodes$S, model$nodes$V, tabulate(nid[vacc] + 1L, model$nodes$count), model$tick)
}
wane <- function(model) {                  # the new piece: V -> S on timer expiry
  V <- model$states[["V"]]; S <- model$states[["S"]]
  w <- step_timer_expire(model$people$state, model$people$timer, model$people$nodeid,
                         model$people$count, model$nodes$count, V, S)
  move_count(model$nodes$V, model$nodes$S, w, model$tick)
}
vaxw <- run_model(data.frame(population = rep(pop, n_nodes), I = rep(50L, n_nodes)),
                  "SIR", nticks = nticks, r0 = 2, infectious_period = dist_gamma(2, 4),
                  network = matrix(0, n_nodes, n_nodes), seed = 1L, extra_states = "V",
                  step_update = wane, step_exit = campaign_timed)
par(mfrow = c(1, 1), mar = c(4, 4.5, 2.5, 1))
matplot(yr, vaxw$nodes$V$values() / 1e3, type = "l", lty = 1, lwd = 2, col = col,
        xlab = "year", ylab = "vaccinated (000s)",
        main = "With waning, V sawtooths (campaign jumps, then decay) — campaigns must repeat")
legend("topleft", c("node 0", "node 1", "node 2 (control)"), col = col, lwd = 2, bty = "n")
```

Permanent `V` *staircased* up; waning `V` *sawtooths* — the same campaign code, plus one
timer and one `step_update`. This composability is the point.

## 3. Quarantine (a "Q" state with a different timed transition)

The same recipe gives case isolation. A periodic, **leaky** testing programme detects each
infectious agent with probability `sensitivity` and moves it `I → Q`; isolated agents stop
transmitting (`Q` isn't in the `I` census). After a fixed period a `step_update` releases
them `Q → R` — here `step_timer_expire` targets `R` instead of `S`.

```{r quarantine}
test_ticks <- seq(0L, 199L, by = 14L); sensitivity <- 0.8; iso_days <- 10L
detect <- function(model) {
  if (!(model$tick %in% test_ticks)) return(invisible())
  I <- model$states[["I"]]; Q <- model$states[["Q"]]
  s <- model$people$state$values(); nid <- model$people$nodeid$values(); cnt <- model$people$count
  inf <- which(s[seq_len(cnt)] == I); det <- inf[runif(length(inf)) < sensitivity]   # 20% missed
  if (!length(det)) return(invisible())
  s[det] <- Q; tm <- model$people$timer$values(); tm[det] <- iso_days
  model$people$state$set(s); model$people$timer$set(tm)
  move_count(model$nodes$I, model$nodes$Q, tabulate(nid[det] + 1L, model$nodes$count), model$tick)
}
release <- function(model) {
  Q <- model$states[["Q"]]; R <- model$states[["R"]]
  r <- step_timer_expire(model$people$state, model$people$timer, model$people$nodeid,
                         model$people$count, model$nodes$count, Q, R)
  move_count(model$nodes$Q, model$nodes$R, r, model$tick)
}
base <- run_model(data.frame(population = 5e5L, I = 100L), "SIR", nticks = 200L,
                  r0 = 2.5, infectious_period = dist_gamma(2, 4), seed = 1L)
quar <- run_model(data.frame(population = 5e5L, I = 100L), "SIR", nticks = 200L,
                  r0 = 2.5, infectious_period = dist_gamma(2, 4), seed = 1L,
                  extra_states = "Q", step_update = release, step_exit = detect)
matplot(0:199, cbind(baseline = rowSums(base$nodes$I$values()),
                     quarantine = rowSums(quar$nodes$I$values())) / 1e3,
        type = "l", lty = 1, lwd = 2.5, col = c("#d7301f", "#2c7fb8"),
        xlab = "day", ylab = "infectious (000s)",
        main = "Fortnightly 80%-sensitive case isolation flattens the curve")
legend("topright", c("baseline", "with quarantine"), col = c("#d7301f", "#2c7fb8"), lwd = 2.5, bty = "n")
```

## Extend further

The same three ingredients — a registered state, a callback that re-labels agents +
`move_count`s the census, and `step_timer_expire` for any timed return — compose into much
more:

- **Stack interventions.** Register `c("V", "Q")` together and run vaccination *and*
  isolation callbacks in the same model.
- **Other timed transitions.** `step_timer_expire(from, to)` works for any pair — a
  treatment state `T → R`, a hospitalised `H → R`/`H → D`, post-infection temporary
  immunity, etc.
- **Targeting by attributes.** Add a per-agent Column in `init` (age, risk group, a
  `dob`) and select intervention recipients with `bincount_where()` predicates (e.g.
  vaccinate only under-fives: `dob > tick - 5*365`).
- **Reactive triggers.** Make a campaign fire when incidence crosses a threshold by testing
  `model$nodes$incidence` inside `step_enter` instead of using a fixed schedule.
- **Drivers.** `step_update` runs just before `calc_foi`, so edit `model$nodes$beta` /
  `model$network` there to model behaviour change or lockdowns affecting *this* tick's
  transmission (census edits, by contrast, affect the next tick — see `?run_model`).
