---
title: 'Using and extending Razer: from R callbacks to custom Rust kernels'
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 4.5, message = FALSE, warning = FALSE)
library(razer)
```

> **Who this is for.** Human modelers *and* coding assistants (LLMs). It is the single
> reference for *how to use* Razer and *how to extend it* — and, critically, where the line
> sits between what you can do **purely in R** and what is better done by **writing a new
> Rust kernel**. The last two sections are a setup checklist and a copy-paste prompt for
> getting an LLM to write a correct Rust kernel for you.
>
> **For LLMs/agents:** the hard rules are collected as a checklist in §7. The repository's
> machine-readable conventions live in `CLAUDE.md`; the kernel exemplars are
> `src/rust/src/steps.rs` and `src/rust/src/transmission.rs`.

There are **three levels** of working with Razer, in increasing power and effort:

| Level | You write… | Use it when… |
|---|---|---|
| **1. Run a model** | one `run_model()` call | your model is in the SI…SEIRS menagerie |
| **2. Customize in R** | R callbacks + composed kernels | you add states, interventions, vital dynamics, custom reports |
| **3. Extend in Rust** | a new `#[extendr]` kernel | you need a *novel per-agent computation* that no existing kernel expresses |

Most users never leave Level 2. The whole point of this guide is to help you stay there as
long as possible, and to make Level 3 approachable (by you or an LLM) when you genuinely
need it.

---

# 1. Level 1 — run a model

`run_model()` builds the agent population and per-node census from a `scenario`
data frame, runs any of the eight closed-population models in the correct per-tick order,
and returns a **`model` environment**.

```{r basic}
m <- run_model(
  data.frame(population = 1e5, I = 100L),   # one node; seed 100 infectious
  model = "SIR", nticks = 120L, r0 = 2.5,
  infectious_period = 8, seed = 1L
)
S <- m$nodes$S$values()[, 1]; I <- m$nodes$I$values()[, 1]; R <- m$nodes$R$values()[, 1]
matplot(0:119, cbind(S, I, R), type = "l", lty = 1, lwd = 2,
        col = c("#2c7fb8", "#d7301f", "#238b45"), xlab = "day", ylab = "agents")
legend("right", c("S", "I", "R"), col = c("#2c7fb8", "#d7301f", "#238b45"), lwd = 2, bty = "n")
```

Key objects on the returned `model`:

- `m$people` — the agent arrays (`Column`s): `state` (`u8`), `timer` (`u16`), `nodeid`
  (`u16`), plus `count`/`capacity`. Agent data lives in Rust; read a snapshot with
  `$values()`.
- `m$nodes` — per-node, per-tick census (`S`/`E`/`I`/`R`/…) and flow (`incidence`,
  `onset`, `recovery`, `waning`) report `Column`s, each an `n_ticks × n_nodes` matrix via
  `$values()`.
- `m$network`, `m$carry`, `m$states` (the state→code map), `m$tick` (current interval).

The model menagerie and the per-tick ordering are covered in the
[final-size](attack_fraction.html) and [endemic dynamics](endemic_dynamics.html) articles.
Everything below is about going *beyond* a single `run_model()` call.

---

# 2. Level 2 — customize **purely in R**

This is where most extension happens, and it requires **no Rust and no recompilation**. You
have two levers: **callbacks** and **extra states**, composed from a toolbox of
R-callable kernels.

## 2.1 The callback lifecycle

`run_model()` accepts an `init` callback (run once, before the loop) and three per-tick
callbacks. The per-tick order is fixed:

```
carry_forward → step_enter → step_*(kernel) → step_update → calc_foi → transmission → step_exit
```

| Callback | Runs | Use it to… |
|---|---|---|
| `init(model)` | once, before tick 0 | allocate extra `Column`s (e.g. `dob`/`dod`), build rate grids |
| `step_enter(model)` | start of each tick | per-tick bookkeeping before dynamics |
| `step_update(model)` | after the step kernel, **before `calc_foi`** | edit drivers (`beta`/`seasonality`/`network`) to affect *this* tick; run extra timed transitions |
| `step_exit(model)` | end of each tick | births/deaths/imports, interventions, custom reports |

**Timing subtlety (read this):** `calc_foi` reads the *settled* census `I[t]`. So a **census
edit** in `step_update`/`step_exit` affects the **next** tick's force of infection, while a
**driver edit** (`beta`, `seasonality`, `network`) in `step_update` affects the **current**
tick. `model$tick` is the current 0-based interval index.

## 2.2 Add agent states with `extra_states`

`extra_states` registers states beyond the model's own S/E/I/R:

- A **known** name — `"M"` (maternal immunity) — keeps its built-in code, and `run_model`
  applies the kernels' M→S waning for you (recording the `waning_m` flow).
- A **new** name — e.g. `"V"` (vaccinated), `"Q"` (quarantined) — is assigned the next free
  `u8` code and becomes a genuine agent state. **The disease kernels ignore unknown states**
  (an agent in `V` is never infected or transitioned), so *you* drive its transitions from a
  callback. The codes are exposed in `model$states`.

```r
# A vaccinated state, driven entirely from R — no Rust needed.
m <- run_model(scenario, "SIR", nticks = 365L, r0 = 2,
               infectious_period = 8, extra_states = "V", seed = 1L,
               step_exit = function(model) {
                 if (model$tick == 30L) {                       # one-time campaign on day 30
                   # move some S -> V at 70% coverage in node 0 (illustrative)
                   # (use a kernel/report to pick agents; apply the count to the census)
                   move_count(model$nodes$S, model$nodes$V, vaccinated_per_node, model$tick)
                 }
               })
```

## 2.3 The R-callable kernel toolbox

These compiled kernels are callable from R and are the building blocks you compose in
callbacks. **Each agent-loop kernel mutates the agent arrays and *returns per-node
counts*;** you apply those counts to the census with `move_count(from, to, counts, tick)`
(`from`/`to` may be `NULL` for one-sided moves — a death decrements only, a birth increments
only).

| Kernel | Does | Returns |
|---|---|---|
| `move_count(from, to, counts, tick)` | apply a transition's per-node counts to the census | — |
| `step_timer_expire(state, timer, nodeid, count, n_nodes, from, to)` | generic timed transition into an **untimed** state (e.g. `V→S` waning, `Q→R`) | per-node counts |
| `step_timer_expire_set(…, from, to, duration)` | generic timed transition into a **timed** state (sets a fresh timer) | per-node counts |
| `births(…)` / `mortality(…)` | vital dynamics (newborns into `M`; retire by date-of-death) | counts |
| `import_infections(…)` | schedule-driven external seeding | counts |
| `constant_pop_vitals_sir(…)` | constant-population turnover | counts |
| `squash(people)` | compact dead agents out, reclaim slots | new live count |
| `bincount` / `bincount_wt` / `bincount_where` / `bincount_where_wt` | roll a per-agent property up into per-node counts (filtered/weighted) | integer vector / writes a report slice |
| `values_map(value, n_ticks, n_nodes)` | broadcast a scalar/vector/matrix into a driver grid | a `Column` |

**Worked patterns** (all pure R, all in the [interventions](interventions.html) and
[vital dynamics](vital_dynamics_measles.html) articles):

- **Vaccination campaign** — `extra_states = "V"`, move `S→V` in a `step_exit` callback.
- **Waning vaccine** — additionally run `step_timer_expire(V → S)` in `step_update`.
- **Quarantine** — `extra_states = "Q"`, detect infectious into `Q` (a leaky, scheduled test),
  release `Q→R` with `step_timer_expire`.
- **Vital dynamics** — `births`/`mortality` in `step_exit`; size with `calc_capacity*`.
- **Importation** — `import_infections` in `step_exit` (needs `capacity` headroom).
- **Custom reports** — `bincount_where(nodeid, n_nodes, dob, "gt", tick - 5*365, count)` for
  under-fives-by-node; write into an allocated report `Column`.

## 2.4 What you can do **purely in R** — checklist

If your idea is on this list, **do not write Rust**:

- ✅ Any of the eight menagerie models, on one node or a spatial network.
- ✅ Add new agent states and drive their transitions from callbacks.
- ✅ Interventions: vaccination (± waning), quarantine/isolation, treatment, ring strategies —
  anything expressible as *moving agents between states* using the kernels above.
- ✅ Vital dynamics, importation, age targeting, time-/space-varying `beta`/`seasonality`.
- ✅ Custom per-node reports via the `bincount` family and your own report `Column`s.
- ✅ Anything that is a **composition of existing timed transitions + per-tick R glue** that
  operates on *whole columns* or on *returned per-node counts* (not on individual agents in
  an R loop).

---

# 3. The R-vs-Rust decision boundary

The single question that decides whether you need Rust:

> **Does your new behavior require a novel computation *per agent, every tick*, that is not a
> composition of existing kernels?**

- **No → stay in R.** Compose `step_timer_expire(_set)`, `move_count`, the `bincount` family,
  and the vital/import kernels in callbacks. These already contain the fast per-agent loops;
  your R code only orchestrates them and edits whole `Column`s or per-node count vectors.
- **Yes → write a Rust kernel.** R cannot run a fast per-agent inner loop (no JIT; an R
  `for` over millions of agents is orders of magnitude too slow), and you cannot mutate the
  Rust-owned `Column`s element-by-element from R without copying them out and back (which
  defeats the design and is usually infeasible for `u8`/`u16` in-place updates).

**Concrete examples**

| Want to… | R or Rust? | Why |
|---|---|---|
| Vaccinate a fraction of S each campaign | **R** | a state move; `move_count` |
| Wane V→S after a timer | **R** | `step_timer_expire` |
| Quarantine detected infectious for 10 days | **R** | extra state + `step_timer_expire` |
| Targeted reports (under-5 by node) | **R** | `bincount_where` |
| A **bespoke force-of-infection** formula (not `calc_foi`'s) | **Rust** | a new per-agent/per-node inner loop |
| **Within-host** progression with per-agent heterogeneity not expressible as a fixed timed chain | **Rust** | novel per-agent update each tick |
| **Agent-to-agent** contact/pairing (households, networks of individuals) | **Rust** | inner loop over agent pairs |
| A new per-agent property updated every tick with custom logic (e.g. waning antibody titre → susceptibility) | **Rust** | per-agent computation over millions of agents |

A useful tell: if you find yourself wanting to write `for (i in seq_len(people$count))` in R,
or to read a whole agent `Column` into R, transform it, and write it back **every tick** —
that is the signal to move the loop into a Rust kernel.

---

# 4. Level 3 — set up for Rust development

Razer's kernels are Rust, compiled at install via [extendr](https://extendr.github.io/).
To add one you need a Rust toolchain and the package's dev tooling.

## 4.1 One-time setup

```bash
# 1. Install the Rust toolchain (rustc + cargo) via rustup:
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh

# 2. cargo installs to ~/.cargo/bin, which is often NOT on the default PATH.
#    Make sure it is on PATH for R's compile/document steps:
export PATH="$HOME/.cargo/bin:$PATH"
```

```r
# 3. R-side dev tooling:
install.packages(c("devtools", "rextendr", "testthat"))
```

Razer's `DESCRIPTION` already declares `SystemRequirements: Cargo … rustc >= 1.65.0`.

## 4.2 The edit → build → test loop

```bash
cd src/rust && cargo check          # fast compile-only check while iterating
```

```r
devtools::document()   # compiles the Rust, regenerates R/extendr-wrappers.R + man/*.Rd
devtools::load_all()   # load the freshly built package
devtools::test()       # run the testthat suite
```

Always use **`devtools::document()`** after changing a Rust signature (it recompiles *and*
regenerates the R wrappers and help pages). Do **not** use `rextendr::document()` (deprecated).
Rust **panics** (`assert!`, `panic!`, index out of bounds) are caught at the extendr boundary
and surface in R as `stop()` errors — so `assert!` is the idiomatic way to validate inputs.

---

# 5. Level 3 — the kernel contract (ABI)

Every Razer agent-loop kernel follows the same shape. Read
`src/rust/src/steps.rs` (the cleanest exemplar) and `src/rust/src/transmission.rs` alongside
this.

**The conventions, as hard rules:**

1. **Signature.** Take the agent `Column`s you read/mutate, an `i32 count` (active agents),
   an `i32 n_nodes`, and any `&Distribution`s. Take `Column`s by `&` (read) or `&mut` (write).
2. **Storage types.** `state` is `u8`, `timer` is `u16`, `nodeid` is `u16`. Get a typed slice
   with the matching accessor: `state.as_u8_mut()`, `timer.as_u16_mut()`, `nodeid.as_u16()`,
   and (`as_u32`/`as_i32_mut`/`as_f64`/…) for other dtypes. These panic on a dtype mismatch.
   Always slice to the live prefix `[..count]`.
3. **Return per-node counts; touch no census buffers.** Return a `List` of
   `integer[n_nodes]` (named, e.g. `list!(waned = …, onset = …)`) or a single
   `Vec<i32>`. The R caller applies them with `move_count`. *Never* write the node census
   inside the kernel.
4. **Parallelize with per-node tallies.** Use `par_chunks_mut(rng::RNG_CHUNK)`, accumulate a
   private `vec![0i64; k * n]` per chunk, and `reduce` by summing. No shared writes.
5. **Randomness goes through `rng`.** Pull one `let base = rng::next_call_base();` *before*
   the parallel fan-out, then `let mut rng = rng::chunk_rng(base, chunk_index);` inside each
   chunk. This is what makes a seeded run reproducible **independent of thread count** — do
   not use `thread_rng()` or seed per-thread.
6. **Decrement guard.** `if t[j] > 0 { t[j] -= 1; }` before testing `== 0` (u16 must not
   underflow), and branch on the agent's state so a just-arrived state is not advanced twice.
7. **Validate with `assert!`** (becomes an R error).
8. **Register it.** Add `fn your_kernel;` to the file's `extendr_module! { … }`, and ensure
   the file's module is listed in `src/rust/src/lib.rs` (both `mod x;` and `use x;` inside its
   `extendr_module!`). If it's a new file, add both lines.

## 5.1 A minimal annotated skeleton

A generic kernel that flips agents in `from_state` to `to_state` with per-agent probability
`p` and returns per-node counts — the template most new kernels start from:

```rust
// src/rust/src/myrule.rs
use extendr_api::prelude::*;
use rayon::prelude::*;
use crate::column::Column;
use crate::rng;
use rand::Rng;                                              // for rng.gen::<f64>()

/// Flip each agent in `from_state` to `to_state` with probability `p`.
///
/// @param state   Per-agent u8 state Column (mutated).
/// @param nodeid  Per-agent u16 node-id Column.
/// @param count   Number of active agents.
/// @param n_nodes Number of nodes (length of the returned vector).
/// @param from_state,to_state  u8 state codes (as i32).
/// @param p       Per-agent flip probability in [0, 1].
/// @return Integer vector of per-node flip counts.
/// @export
#[extendr]
fn flip_state(
    state: &mut Column,
    nodeid: &Column,
    count: i32,
    n_nodes: i32,
    from_state: i32,
    to_state: i32,
    p: f64,
) -> Vec<i32> {
    assert!(count >= 0 && n_nodes >= 0, "`count`/`n_nodes` must be non-negative");
    assert!((0.0..=1.0).contains(&p), "`p` must be in [0, 1]");
    let (count, n) = (count as usize, n_nodes as usize);
    let (from, to) = (from_state as u8, to_state as u8);

    let base = rng::next_call_base();                       // ONE base per call, before fan-out
    let st  = &mut state.as_u8_mut()[..count];
    let nid = &nodeid.as_u16()[..count];

    let tally: Vec<i64> = st
        .par_chunks_mut(rng::RNG_CHUNK)
        .enumerate()
        .zip(nid.par_chunks(rng::RNG_CHUNK))
        .map(|((ci, s), node)| {
            let mut rng = rng::chunk_rng(base, ci);          // deterministic per-chunk stream
            let mut local = vec![0i64; n];
            for j in 0..s.len() {
                if s[j] == from && rng.gen::<f64>() < p {    // rand's Rng::gen: uniform [0, 1)
                    s[j] = to;
                    local[node[j] as usize] += 1;
                }
            }
            local
        })
        .reduce(|| vec![0i64; n], |mut a, b| { for k in 0..n { a[k] += b[k]; } a });

    tally.iter().map(|&x| x as i32).collect()
}

extendr_module! {
    mod myrule;
    fn flip_state;
}
```

…then add `mod myrule;` and `use myrule;` to `lib.rs`, run `devtools::document()`, and call it
from a callback, applying its counts with `move_count`:

```r
step_exit = function(model) {
  flips <- flip_state(model$people$state, model$people$nodeid, model$people$count,
                      length(model$states) - 1L, model$states[["S"]], model$states[["V"]], 0.01)
  move_count(model$nodes$S, model$nodes$V, flips, model$tick)
}
```

> `rng.gen::<f64>()` is `rand`'s `Rng`-trait method for a uniform draw in `[0, 1)` (hence the
> `use rand::Rng;`) — the same per-agent draw `transmission.rs` uses to turn a per-node
> probability into infections.

## 5.2 Don't forget the test

Every kernel ships a test (`tests/testthat/test-*.R`). The standard pattern is a
**parallel-vs-serial census check**: run the kernel on a large population, then confirm the
returned per-node counts equal a direct R tally of the resulting agent states. Reproducibility
is checked by running twice under `set_seed()` and asserting identical results.

---

# 6. Prompting an LLM to write the Rust kernel

An LLM that is given the conventions and an exemplar can write a correct Razer kernel
reliably. The recipe:

**Give it this context:**

1. The **exemplar files** verbatim: `src/rust/src/steps.rs` (the ABI in action) and the
   target file you're adding to. Tell it these are the style/ABI to match.
2. The **eight hard rules** from §5 (return per-node counts, `u16` timers + decrement guard,
   `rng::next_call_base` once + `rng::chunk_rng` per chunk for thread-independent
   reproducibility, `assert!` for validation, register in the file's `extendr_module!` and in
   `lib.rs`).
3. The **commenting convention**: this codebase is read by C/C++/C#/Python programmers who do
   *not* know Rust — comment Rust-specific idioms (`&`/`&mut`, slices, `match`, closures,
   iterators, `Option`/`Result`/`?`, `unsafe`, what `#[extendr]` generates).
4. The **dev loop**: `devtools::document()` then `devtools::test()`; panics become R errors.

**A copy-paste prompt template:**

> You are adding a kernel to **Razer**, an R package whose performance kernels are Rust
> exposed to R via `#[extendr]`. Follow the existing ABI exactly — here is the reference
> exemplar:
>
> ```
> «paste the full contents of src/rust/src/steps.rs»
> ```
>
> **Task:** write a kernel `«name»` that, for each active agent, «describe the per-agent
> computation precisely: which Columns it reads/writes, the condition, the state change,
> any timer/distribution, what per-node quantity to count».
>
> **Requirements (do not deviate):**
> - Signature takes the agent `Column`s, `count: i32`, `n_nodes: i32`, and any
>   `&Distribution`s; take `Column`s by `&`/`&mut`; slice to `[..count]`.
> - `state` is `u8`, `timer` is `u16`, `nodeid` is `u16`; use `as_u8_mut`/`as_u16_mut`/
>   `as_u16`. Guard timer decrements (`if t>0 { t-=1 }`).
> - **Return per-node counts** as `Vec<i32>` or a named `List` of `integer[n_nodes]`; do NOT
>   write any node census inside the kernel.
> - Parallelize with `par_chunks_mut(rng::RNG_CHUNK)` + a private per-node `Vec<i64>` reduced
>   by summing.
> - For randomness: `let base = rng::next_call_base();` once before the fan-out, then
>   `rng::chunk_rng(base, chunk_index)` inside each chunk (this keeps seeded runs reproducible
>   regardless of thread count). Never use `thread_rng`.
> - Validate inputs with `assert!`.
> - Add `fn «name»;` to the file's `extendr_module!`, and tell me the exact `lib.rs` edits.
> - Comment Rust idioms for a C/C#/Python reader (per the repo convention).
> - Write a `roxygen`/extendr doc comment with `@param`/`@return`/`@export`.
> - Also write a `testthat` test asserting the returned per-node counts equal a direct R
>   census of the mutated agent states at ~1e6 agents, plus a reproducibility check under
>   `set_seed()`.
>
> After I paste back any compiler errors from `devtools::document()`, fix them.

**Iterate** by pasting compiler/test output back to the LLM. Because panics become R errors
and the test compares against an independent R census, you get a tight correctness loop.

---

# 7. Quick reference for LLMs / agents

Collected hard rules (see §5 for detail):

- **Stay in R** unless the task needs a *novel per-agent computation every tick* that isn't a
  composition of `step_timer_expire(_set)` + `move_count` + the vital/import/`bincount`
  kernels. A wish to write a per-agent `for` loop in R, or to round-trip a whole agent
  `Column` through R each tick, means: write Rust.
- **Kernel ABI:** `#[extendr] fn`, args = agent `Column`s + `count: i32` + `n_nodes: i32` +
  `&Distribution`s; `Column`s by `&`/`&mut`, sliced `[..count]`; `state` `u8` / `timer` `u16`
  / `nodeid` `u16` via `as_u8_mut`/`as_u16_mut`/`as_u16`; **return per-node counts** (a
  `Vec<i32>` or named `List`), never touch the node census; parallelize with
  `par_chunks_mut(rng::RNG_CHUNK)` + per-chunk `Vec<i64>` reduced by sum.
- **RNG:** `rng::next_call_base()` once before fan-out, `rng::chunk_rng(base, ci)` per chunk —
  thread-count-independent reproducibility. Never `thread_rng`.
- **Validate** with `assert!` (→ R error). **Guard** `u16` timer decrements.
- **Register** in the file's `extendr_module!` and in `src/rust/src/lib.rs` (`mod`/`use`).
- **Build:** `devtools::document()` (recompiles + regenerates wrappers/man), then
  `devtools::test()`. Comment Rust idioms for a non-Rust (C/C#/Python) reader.
- **Exemplars:** `src/rust/src/steps.rs`, `src/rust/src/transmission.rs`. Conventions:
  `CLAUDE.md`.
