---
title: 'Long runs and memory: capacity and squash()'
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, message = FALSE, warning = FALSE)
library(razer)
```

> Companion to [`examples/long_run_squash.R`](https://github.com/clorton/razer/blob/main/examples/long_run_squash.R). (The run chunk simulates a century — give it
> ~30 seconds.)

## The problem: agents accumulate

razer is **agent-based** — every individual occupies a slot in fixed-size `Column` arrays.
A closed population is easy: allocate exactly $N$ slots. But once you add **births**, the
array must hold every agent that will *ever exist*. Over a century at a crude birth rate of
30/1,000/year, that is tens of millions of slots even though only a few million are ever
*alive at once*. Allocating for the cumulative total is wasteful; the fix is to **reclaim
the slots of the dead** and reuse them.

Two helpers bracket the choice:

- **`calc_capacity(birthrates, pop)`** — the *cumulative-births* bound: how many agents are
  ever born (with a safety factor). Use it when you **never** reclaim.
- **`calc_capacity_cdr(birthrates, deathrates, pop)`** — the *peak-living* bound: net births
  minus a conservatively under-counted death rate. Use it when you reclaim dead slots
  periodically with **`squash()`**.

`squash(people)` compacts the living agents to the front of every per-agent Column (keeping
them aligned) and drops the dead, returning the new live count — so the active count tracks
the *living* population instead of the cumulative total.

```{r capacity}
N0 <- 500000L; years <- 100L; nticks <- years * 365L; cbr <- 30; cdr <- 15
br <- matrix(cbr, nticks - 1L, 1L); dr <- matrix(cdr, nticks - 1L, 1L)
cap_cdr   <- as.integer(calc_capacity_cdr(br, dr, N0, safety_factor = 1))   # we allocate this
cap_naive <- as.integer(min(calc_capacity(br, N0, safety_factor = 1), .Machine$integer.max))
cat(sprintf("calc_capacity_cdr (peak-living, with squash) = %s slots\n", format(cap_cdr, big.mark = ",")))
cat(sprintf("calc_capacity     (cumulative, no reclaim)   = %s slots  (%.1fx more)\n",
            format(cap_naive, big.mark = ","), cap_naive / cap_cdr))
```

## A century with annual squash

No disease here (model `"SI"` with `r0 = 0`, so nothing transmits) — just demography. Crude,
age-independent mortality is an exponential lifespan at the CDR hazard (a Kaplan–Meier table
built from an exponential survival curve). Newborns enter the maternal state `M` (registered
via `extra_states`; `run_model` wanes M→S for us). The `step_exit` callback runs `births`
and `mortality` each tick and `squash()`es once a year, recording the living and active
counts for the plots.

```{r run}
h_year <- -log(1 - cdr / 1000)                                  # constant per-year hazard
km <- kaplan_meier_estimator(round((1 - exp(-h_year * (0:200))) * 1e7))   # exponential life table
mat_dur <- dist_normal(270, 20); squash_every <- 365L

track <- new.env()
track$live <- numeric(nticks); track$live[1] <- N0
track$active <- numeric(nticks); track$active[1] <- N0

m <- run_model(data.frame(population = N0), "SI", nticks = nticks, r0 = 0,
               infectious_period = dist_constant(7),            # unused (r0 = 0)
               capacity = cap_cdr, extra_states = "M", seed = 1L,
               init = function(model) {
                 cap <- model$people$capacity
                 model$people$dob <- allocate_scalar("i32", cap)
                 model$people$dod <- allocate_scalar("u32", cap)
                 model$nodes$birth_rate <- values_map(cbr / 1000 / 365, nticks, 1L)
                 age <- pmin(as.integer(floor(stats::rexp(N0, h_year / 365))), 90L * 365L)
                 model$people$dob$set(c(-age, integer(cap - N0)))
                 model$people$dod$set(c(as.integer(km$predict_age_at_death(age, -1L) - age),
                                        integer(cap - N0)))
               },
               step_exit = function(model) {
                 t <- model$tick; idx <- t + 2L
                 b <- births(model$people$state, model$people$timer, model$people$nodeid,
                             model$people$dob, model$people$dod, model$people$count, 1L,
                             model$nodes$birth_rate, mat_dur, km, t)
                 model$people$count <- b$count; move_count(NULL, model$nodes$M, b$born, t)
                 d <- mortality(model$people$state, model$people$dod, model$people$nodeid,
                                model$people$count, 1L, t)
                 move_count(model$nodes$M, NULL, d$m, t); move_count(model$nodes$S, NULL, d$s, t)
                 if ((t + 1L) %% squash_every == 0L) model$people$count <- squash(model$people)
                 track$live[idx]   <- sum(model$nodes$M$values()[idx, ], model$nodes$S$values()[idx, ])
                 track$active[idx] <- model$people$count
               })
cat(sprintf("living %s -> %s (%.1fx growth); peak active = %s of %s capacity (%.0f%% used)\n",
            format(N0, big.mark = ","), format(round(track$live[nticks]), big.mark = ","),
            track$live[nticks] / N0, format(round(max(track$active)), big.mark = ","),
            format(cap_cdr, big.mark = ","), 100 * max(track$active) / cap_cdr))
```

```{r plots}
yr <- (seq_len(nticks) - 1L) / 365
par(mfrow = c(2, 1), mar = c(4, 4.5, 2.5, 1))
plot(yr, track$live / 1e6, type = "l", lwd = 2.5, col = "#238b45", ylim = c(0, cap_naive / 1e6 * 1.05),
     xlab = "year", ylab = "agents (millions)", main = "Living population vs the two capacity bounds")
abline(h = cap_cdr / 1e6, lty = 2, lwd = 2, col = "#d7301f")
abline(h = cap_naive / 1e6, lty = 3, lwd = 2, col = "grey40")
legend("topleft", c("living population", "calc_capacity_cdr (allocated)", "calc_capacity (no reclaim)"),
       col = c("#238b45", "#d7301f", "grey40"), lwd = c(2.5, 2, 2), lty = c(1, 2, 3), bty = "n")
matplot(yr, cbind(active = track$active, living = track$live) / 1e6, type = "l", lty = 1, lwd = 2,
        col = c("#2c7fb8", "#238b45"), ylim = c(0, cap_cdr / 1e6 * 1.05),
        xlab = "year", ylab = "agents (millions)", main = "Annual squash() keeps active slots near the living count")
abline(h = cap_cdr / 1e6, lty = 2, lwd = 2, col = "#d7301f")
legend("topleft", c("active slots (people$count)", "living population", "allocated capacity"),
       col = c("#2c7fb8", "#238b45", "#d7301f"), lwd = c(2, 2, 2), lty = c(1, 1, 2), bty = "n")
```

The living population grows steadily, and the **active slot count tracks it closely** —
staying well under the allocated `calc_capacity_cdr` line and far below the
`calc_capacity` (no-reclaim) bound. That gap is the memory `squash()` saves: without it you
would have to allocate for every agent ever born.

## Customize and extend

- **Squash cadence.** Squash less often (e.g. every 5 years) and the active count sawtooths
  higher between reclaims (dead slots accumulate, then drop) — trading a little head-room
  for fewer compactions. Size `capacity` for the worst-case inter-squash peak.
- **Safety factor.** `calc_capacity_cdr(..., safety_factor = s)` credits only `1/(1+s)` of
  the deaths, so larger `s` reserves more slack against a lower-mortality (faster-growing)
  realization; `s = 0` is the bare net-growth bound.
- **Add disease.** Swap `"SI", r0 = 0` for a real model with `incubation_period` /
  transmission and you have an endemic, demographically open simulation that can run for
  centuries within a bounded array (this is exactly what a long measles or polio run needs).
