---
title: 'Spatial epidemics: metapopulations and migration networks'
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, message = FALSE, warning = FALSE)
library(razer)
```

> Companion to [`examples/simple_sir.R`](https://github.com/clorton/razer/blob/main/examples/simple_sir.R).

## Why space?

A single well-mixed population is a useful fiction, but real epidemics play out across a
**metapopulation** — a set of patches (towns, districts, countries) each with its own local
transmission, *coupled* by the movement of people. Coupling drives the phenomena a
single-patch model can't show: spatial **invasion waves**, **rescue** of a faded-out patch
by its neighbours, and partial **synchrony** between patches.

Razer models a metapopulation as $n$ patches each running the usual SIR transmission, plus a
coupling matrix $W$ where $W_{ij}$ is the fraction of patch $i$'s force of infection that is
*exported* to patch $j$. The kernel `calc_foi` redistributes each tick:

$$\lambda_k \;=\; r_k\Big(1 - \textstyle\sum_j W_{kj}\Big) \;+\; \sum_i r_i\,W_{ik}, \qquad r_k = \beta_k\,\frac{I_k}{N_k},$$

i.e. patch $k$ keeps the un-exported part of its own local force and imports a share of
every other patch's.

## Building a coupling network

We use the England & Wales measles patches (954 registration districts, each with a 1944
population and a latitude/longitude). Razer ships several **migration models** that turn
pairwise distances + populations into a weight matrix — `gravity()`, `radiation()`,
`stouffer()`, `competing_destinations()` — plus `row_normalizer()` to cap each patch's total
emigration so the rows are valid export fractions. Here: the parameter-free **radiation
model** (Simini et al., *Nature* 2012), capped at 10% emigration per patch.

```{r network}
# A pkgdown article under vignettes/articles/, so the shared example data is two up.
scenario <- read.csv(file.path("..", "..", "examples", "data", "EnglandWalesMeasles_places.csv"))
D  <- distances(scenario$latitude, scenario$longitude)            # pairwise great-circle km
W  <- radiation(scenario$population, D, k = 1, include_home = FALSE)
W  <- row_normalizer(W, max_rowsum = 0.1)                         # cap total emigration at 10%
cat(sprintf("%d patches; total population %s; network row-sums in [%.3f, %.3f]\n",
            nrow(scenario), format(sum(scenario$population), big.mark = ","),
            min(rowSums(W)), max(rowSums(W))))
```

## Run a spatial SIR

`run_model()` takes the coupling matrix directly as `network`; internally it is stored once
as a 2-D Column so `calc_foi` reads it Rust-side each tick. We seed a few infectious and
~5% immune in every patch and run a daily SIR for ~4 months.

```{r run}
scenario$I <- pmin(5L, scenario$population)                       # 5 infectious / patch
scenario$R <- pmin(as.integer(0.05 * scenario$population), scenario$population - scenario$I)
m <- run_model(scenario, "SIR", nticks = 120L, r0 = 2,
               infectious_period = dist_gamma(2, 2), network = W, seed = 1L)

days <- 0:119
S <- rowSums(m$nodes$S$values()); I <- rowSums(m$nodes$I$values()); R <- rowSums(m$nodes$R$values())
matplot(days, cbind(S, I, R) / 1e6, type = "l", lty = 1, lwd = 2.5,
        col = c("#2c7fb8", "#d7301f", "#238b45"), xlab = "day", ylab = "agents (millions)",
        main = "National S / I / R, summed over 954 patches")
legend("right", c("S", "I", "R"), col = c("#2c7fb8", "#d7301f", "#238b45"), lwd = 2.5, bty = "n")
```

## Map the burden

Because the model is spatial, the interesting output is *per patch*. Each patch's census is
its own column of `m$nodes$I$values()`; here we map the **attack rate** (fraction ever
infected) at each patch's coordinates, sized by population — the recognizable shape of
England & Wales.

```{r map}
attack <- colSums(m$nodes$incidence$values()) / scenario$population   # per-patch fraction infected
pal <- grDevices::colorRampPalette(c("#ffffb2", "#fd8d3c", "#bd0026"))(64)
rng <- range(attack); idx <- pmax(1L, pmin(64L, 1L + round(63 * (attack - rng[1]) / max(diff(rng), 1e-9))))
plot(scenario$longitude, scenario$latitude, pch = 19, asp = 1, col = pal[idx],
     cex = 0.3 + 1.8 * sqrt(scenario$population / max(scenario$population)),
     xlab = "longitude", ylab = "latitude", main = "Attack rate by patch (size ~ population)")
lv <- pretty(rng, 4)
legend("topleft", title = "attack rate", bty = "n", pch = 19, pt.cex = 1.4,
       legend = sprintf("%.0f%%", 100 * lv),
       col = pal[pmax(1L, pmin(64L, 1L + round(63 * (lv - rng[1]) / max(diff(rng), 1e-9))))])
```

## Customize and extend

- **Migration model.** Swap `radiation()` for `gravity(population, D, ...)` (tunable
  distance decay) or `stouffer()` / `competing_destinations()`; tighten/loosen
  `row_normalizer(max_rowsum = …)` to change how strongly patches couple.
- **Watch an invasion wave.** Seed a *single* patch (e.g. the largest) with `I` and set
  `R = 0` everywhere, then map the per-patch *day of peak* (`apply(I_mat, 2, which.max)`):
  patches light up later the farther they sit from the seed, along the network.
- **Endemic + spatial.** Add the constant-population vitals `step_exit` callback from the
  endemic notebook (and a `capacity` for imports) to keep patches endemic and study spatial
  persistence / rescue effects below the critical community size.
- **Seasonality per patch.** `run_model`'s `seasonality` accepts a full `n_ticks x n_nodes`
  grid (via `values_map`), so different patches can have offset school terms or climates.
